% Metadata for PDF/A
\begin{filecontents*}{\jobname.xmpdata}
\Author{Diego Michele Monti}
\Title{Multicriteria Evaluation for Top-k and Sequence-based Recommender Systems}
\Subject{Doctoral dissertations in the SCUDO doctoral school}
\Keywords{PDF\sep
          PDF/A\sep
          ISO 19005\sep
          LaTeX\sep
          PhD Thesis\sep
          Engineering\sep
          SCUDO}
\Publisher{Politecnico di Torino}
\end{filecontents*}

\documentclass[%
12pt,
twoside,
tipotesi=scudo,
numerazioneromana,
final
]{toptesi}

\usepackage[a-1b]{pdfx}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\errorcontextlines=9

% Fonts
\usepackage{lmodern}

\makeindex[intoc]

% Other packages
\usepackage{enumitem}
\usepackage{listings}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{tabu}
\usepackage{nameref}
\usepackage{tikz}
\usepackage{todonotes}
\usepackage{changes}
\usepackage{tabularx}
\usepackage{algorithm}[chapter]
\usepackage{algorithmic}
\usepackage{rotating}

% Show all authors in bibliography
\AtBeginBibliography{\setcounter{maxnames}{99}}

% SLR
\lstset{basicstyle=\ttfamily,breaklines=true,breakindent=0em,breakatwhitespace=true}

\newcounter{papercounter}
\newcommand{\paper}[1]{\refstepcounter{papercounter}\label{#1}P\arabic{papercounter}}

\newcounter{papervalue}
\newcommand{\pref}[1]{\setcounterref{papervalue}{#1}\hyperref[#1]{[P\arabic{papervalue}]}}

\makeatletter
\let\orgdescriptionlabel\descriptionlabel
\renewcommand*{\descriptionlabel}[1]{%
  \let\orglabel\label
  \let\label\@gobble
  \phantomsection
  \edef\@currentlabel{#1\unskip}%
  \let\label\orglabel
  \orgdescriptionlabel{#1}%
}
\makeatother

% Sequeval
\newtheorem{definition}{Definition}

% RNN
\newcommand{\qu}[1]{``{#1}''}

% Algorithms
\renewcommand{\thealgorithm}{\arabic{chapter}.\arabic{algorithm}}

% Hyperref
\unless\ifcsname ver@hyperref.sty\endcsname\usepackage{hyperref}\fi
\hypersetup{%
    pdfpagemode={UseOutlines},
    bookmarksopen,
    pdfstartview={FitH},
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}

\usepackage{indentfirst}
\raggedbottom

\begin{document}

% First page
\begin{ThesisTitlePage}

\PhDschoolLogo{Logo-ScuDo-blu}
\ProgramName{Computer and Control Engineering}
\CycleNumber{32\textsuperscript{th}}

\author{Diego Michele Monti}
\title{Multicriteria Evaluation for\\Top-k and Sequence-based Recommender Systems}

\NSupervisor{Supervisor:}{Supervisors:}
\SupervisorNumber{2}
\SupervisorList{Prof.~Maurizio Morisio, Supervisor\\
Dr.~Giuseppe Rizzo, Co-supervisor}

\ExaminerList{%
Prof.~Luis Mart\'inez-L\'opez, Referee, Universidad de Ja\'en\\
Prof.~Olga C. Santos, Referee, Universidad Nacional de Educaci\'on a Distancia\\
Prof.~Andrea Bottino, Politecnico di Torino\\
Prof.~Marco Torchiano, Politecnico di Torino\\
Dr.~Iacopo Vagliano, Amsterdam UMC}

\Nlocation{Politecnico di Torino}
\ExaminationDate{May 7\textsuperscript{th}, 2020}

\CClicence{%
\noindent This dissertation is licensed under a Creative Commons Attribution -- NonCommercial -- NoDerivatives 4.0 International License. Visit \url{http://creativecommons.org/licenses/by-nc-nd/4.0/} to view a copy of this license.}

\Disclaimer{%
\noindent I hereby declare that, the contents and organisation of this dissertation constitute my own original work and does not compromise in any way the rights of third parties, including those relating to the security of personal data.	
}

\Signature{%
\begin{flushright}
\parbox{0.5\textwidth}{\centering
%\dotfill\\
Diego Michele Monti\\
Turin, May 7\textsuperscript{th}, 2020
}
\end{flushright}
}

\end{ThesisTitlePage}

\summary

\paragraph{Context}

The need of effectively identifying relevant items from a potentially overwhelming catalog has led to the creation of automatic approaches for supporting users of online platforms in such a time-consuming task. Recommender systems are software tools and algorithms designed to suggest items to users according to their preferences. The traditional goal of the recommendation problem is to create a ranked list of suggestions for each user. However, a more novel paradigm is represented by algorithms capable of exploiting the temporal dimension of the available ratings, known as sequence-based recommender systems. Differently from the field of information retrieval, performing an offline experiment for comparing multiple recommendation approaches is a challenging task, as the ground truth is represented by the subjective preferences of the users collected before the introduction of the system under evaluation. Nevertheless, it represents a powerful tool for selecting the most promising approaches to be further tested in subsequent trails.

\paragraph{Goal}

In this dissertation, we study how to perform the offline evaluation of a generic recommender system by exploiting a multicriteria approach that relies on a set of heterogeneous metrics. We provide an answer to the following research questions: how already established metrics can be extended and adapted for evaluating a sequence-based recommender system; what is the most suitable protocol for performing an offline evaluation of a top-$k$ recommender system; and to what extend the structure of a rating dataset can influence the results of an offline evaluation.

\paragraph{Method}

Regarding the first question, we propose an evaluation framework called Sequeval designed to compare in a replicable way different sequence-based recommenders considering eight dimensions. With respect to the second question, we introduce RecLab, an evaluation toolkit designed following a distributed approach to overcome some limitations of the currently available evaluation solutions. Finally, for addressing the third question, we describe a method based on data visualization to explore the structure of a rating dataset and we report on an algorithm capable of generating alternative versions of an existing collection of ratings. Furthermore, we investigate the topic of multicriteria recommender systems more in general with a systematic literature review, also considering their evaluation protocols.

\paragraph{Results}

From the systematic literature review, we observed that is not possible to directly compare the results obtained with different multicriteria recommendation approaches due to the extreme variability in the reported experimental protocols. We exploited Sequeval and RecLab to successfully conduct some experimental campaigns involving different recommenders and datasets. The availability of multicriteria metrics enables the experimenter to obtain a more comprehensive picture of the systems under consideration. We validated our visualization method and generative approach by qualitatively and quantitatively comparing the results obtained with different rating datasets. Finally, we considered two novel recommendation approaches as possible use cases of multicriteria evaluation methods.

\paragraph{Conclusion}

In summary, this dissertation deals with the problem of conducting an offline comparison of recommender systems considering both traditional and sequential scenarios. We designed the proposed frameworks for addressing the most critical problems that emerged from literature, namely the non-reproducibility of the results, the comparability of different studies, and the bias of relying on a few metrics only. The software code of our frameworks is freely available on the Web in an attempt to foster further reuse and extension.

\acknowledgements

This dissertation represents the end of a journey that lasted about three years and a half and it would not have been possible without the help and the continuous support of different people with whom I had the privilege to collaborate.

First of all, I would like to thank my supervisor, prof. Maurizio Morisio, for encouraging me to pursue a doctorate title and for always helping to face the difficulties that such a choice necessarily implies, and my co-supervisor, Giuseppe Rizzo, for always trying, and hopefully succeeding, in making me a more mature researcher. Indeed, I would like to thank prof. Marco Torchiano, for recruiting me as a teaching assistant since my first year and for introducing me to the topic of data visualization, and Luca Ardito, for the always precious advice and suggestions.

A very special thank goes to Iacopo Vagliano, for helping to orientate myself in the complex world of recommender systems during my early days, and to Enrico Palumbo, for the countless hours spent together debating mathematical equations and configuration parameters of our experiments.

Of course, I would like to mention all the colleagues and friends who contributed to the positive and vibrant atmosphere of Lab 1, strictly in alphabetical order: Alysson Dos Santos, Amirhosein Toosi, Edoardo Battegazzorre, Erion {\c C}ano, Francesco Strada, Isabeau Oliveri, Mariachiara Mecati, Riccardo Coppola, Rifat Rashid, and Simone Leonardi.

Several experimental activities that I conducted would not have been possible without the computational resources kindly provided by the \textsc{hpc@polito} project (\url{http://hpc.polito.it}), to which I am grateful.

Last but not least, I would like to thank my parents for their unconditional love and fundamental support during this journey.

% Index
\tablespagetrue\figurespagetrue
\allcontents

% Glossary
\nomenclature{\textit{collaborative filtering} (CF)}{A recommendation approach based on the rating behaviour of similar users on similar items.}
\nomenclature{\textit{content-based recommender system}}{A recommendation approach based on the correspondence between the preferences of the users and the characteristics of the items.}
\nomenclature{\textit{evaluation framework}}{An evaluation protocol, a set of evaluation metrics, and, if available, a software implementation of them.}
\nomenclature{\textit{evaluation protocol}}{The procedure that needs to be followed by a researcher to conducted an experiment.}
\nomenclature{\textit{hybrid recommender system}}{A recommendation approach that combines different techniques to select the suggested items.}
\nomenclature{\textit{knowledge-based recommender system}}{A recommendation approach based on externally encoded domain knowledge.}
\nomenclature{\textit{multicriteria evaluation}}{An experimental approach based on a comprehensive set of different evaluation metrics.}
\nomenclature{\textit{multicriteria recommender system}}{A recommender system that exploits multiple ratings per item expressed over different criteria.}
\nomenclature{\textit{offline evaluation}}{An experimental approach based on user preferences collected before the introduction of the system under evaluation.}
\nomenclature{\textit{online evaluation}}{An experimental approach that consists in making a novel recommender system available to a potentially large community of users.}
\nomenclature{\textit{rating dataset}}{A dataset containing the preferences expressed by a set of users over a catalog of items at a certain timestamp.}
\nomenclature{\textit{recommender system} (RS)}{A software tool and an algorithm designed to suggest items to users according to their preferences.}
\nomenclature{\textit{sequence-based recommender system}}{A recommender system capable of suggesting personalized sequences of items.}
\nomenclature{\textit{sequential recommender system}}{A recommender system designed to consider the timestamp associated with user ratings.}
\nomenclature{\textit{synthetic dataset}}{A rating dataset created in an artificial way.}
\nomenclature{\textit{top-k recommender system}}{A recommender system capable of suggesting a list of the $k$ most relevant items per user.}
\nomenclature{\textit{user study}}{An experimental approach that consists in making a recommender system available to a limited set of subjects in a controlled environment.}

\renewcommand{\nomname}{Glossary}
\renewcommand\nomgroup[1]{}
\printnomenclature

\mainmatter

\include{Chapter01/chapter01}
\include{Chapter02/chapter02}
\include{Chapter03/chapter03}
\include{Chapter04/chapter04}
\include{Chapter05/chapter05}
\include{Chapter06/chapter06}
\include{Chapter07/chapter07}
\include{Chapter08/chapter08}
\include{Chapter09/chapter09}
\include{Chapter10/chapter10}

\appendix
\include{Appendix01/appendix01}
\include{Appendix02/appendix02}

\backmatter

\sloppy
\printbibliography[heading=bibintoc]

\printindex

\end{document}